---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ati

<!-- badges: start -->
<!-- badges: end -->

The goal of ati is to help student learning about financial machine learning and its application to algorithmic trading and investment.

This package contains templates for reports, and functions and workshops using in [*Algorithmic trading and investment*](https://canvas.qub.ac.uk/courses/11744)) taught by [Barry Quinn](https://quinference.com/) inb Queen's Management School. 

## Installation

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("barryquinn1/ati")
```
## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(ati)
library(tidyverse)
## basic example code
ati::daily_factors %>% summary()
```
## Function test

```{r}
?ati::estRMT()
```


## Tutorials
**The tutorials can be run on a local machine only**.  You can start the tutorials in one of two ways. First, in RStudio 1.3 or later, you will find the ATI tutorials listed in the "Tutorial" tab in the top-right pane (by default). Find a tutorial and click "Run Tutorial" to get started. Second, you can run any tutorial from the R console by typing the following line: 

``` r
learnr::run_tutorial("Workshop2","ati")
```

This should bring up a tutorial in your default web browser. You can see the full list of tutorials by running:

``` r
learnr::run_tutorial(package = "ati")
```

## Critical Essay
This package also includes a RMarkdown template for use in the critical essay assessment.
Go to File>New>R Markdown... and choose from `From Template` then `Report`.

## Datasets
### Top 30 ftse companies
This package also includes monthly returns for the Top 30 FTSE companies.  You can access this data by running the following

``` r
ati::ftse30_returns_mthly
```
### Daily uk asset pricing risk factors

```r
data("daily_factors")
```


